---
title: "CALCOFI- Report"
author: "Afsana Mimi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Summary: 
This report presents a stepwise analysis of the CalCOFI Cast and Bottle dataset to understand and predict ocean temperature in the California Current System. The purpose of the analysis is to build and test a statistical model that predicts ocean temperature from measurements collected from the same water samples —especially salinity, dissolved oxygen, ammonia, and nutrients—across many stations and depths, using all other variables.

# Data Import and Cleaning

## 1: Call Packages

```{r packages}
library(tidyverse)
library(janitor)
library(stringi)
library(rsample)
library(yardstick)
library(broom)
library(ppcor)
library(car)
library(glmnet)
library(Matrix)
library(knitr)
library(readxl)
library(dplyr)
library(stringr)
library(readr)
#install.packages("ggcorrplot")
library(ggcorrplot)

```

## 2. PATH SETUP

```{r path}
# Get the data
data_file <- "C:/Users/afsan/OneDrive/Desktop/Thesis/MERGED_CalCOFI_DATA_CAST_BOTTLE.csv"

```

Since having issue while reading the file, i want to dive dive more to the file, Checking the file name.


I have found out r is reading data not as an example Bottle-pH2, however as it is calling as `Bottle-pH2`.
```{r}

df_raw <- readr::read_csv(
  data_file,
  col_types = readr::cols(.default = readr::col_character()),
  locale = readr::locale(encoding = "Latin1"),
  show_col_types = FALSE
)

head(df_raw)

```
```{r}
# Fix encoding BEFORE cleaning names
names(df_raw) <- stringi::stri_enc_toutf8(names(df_raw))
names(df_raw)

```
```{r}
# Clean names to snake_case
df_raw <- janitor::clean_names(df_raw)

names(df_raw)
```
Read CalCofi DataSets CodeBook Data
```{r}
df_excel <- read_excel("C:/Users/afsan/OneDrive/Desktop/Thesis/CalCofi DataSets CodeBook.xlsx", sheet = "Sheet1")
head(df_excel)
```
```{r}
#Check the names
names(df_excel)
```
```{r}
#Extract Quality Code Field Names
quality_cols <- df_excel %>%
  filter(str_detect(Description, "Quality Code")) %>%
  pull(`Cast\\Bottle Fields`) %>%
  gsub("^\\[|\\]$", "", .) %>%      # remove brackets
  janitor::make_clean_names()       # match df_raw cleaned names

quality_cols
```


```{r}
#Keep Only Rows Where "Quality Code" is Blank OR 4 OR 6 OR 9
df_filtered <- df_raw %>%
  filter(
    if_all(
      all_of(quality_cols),
      ~ is.na(.x) | .x %in% c(4, 6, 9)))

nrow(df_filtered)
nrow(df_raw)
```

```{r}
#Check only quality columns (to confirm filter worked)
df_filtered %>%
  dplyr::select(all_of(quality_cols)) %>%
  head(20)

```
```{r}
#Remove Quality Columns
df_final <- df_filtered %>%
  dplyr::select(-dplyr::all_of(quality_cols))

nrow(df_final)
head(df_final)
```

```{r}
names(df_final)
```

```{r}
# Get rid of temperature-related quality fields
df_final <- df_final %>%
  dplyr::select(
    -bottle_rec_ind,
    -bottle_t_prec
  )

# Confirm they are gone
c("bottle_rec_ind", "bottle_t_prec") %in% names(df_final)

```
# Predictive Analysis Process
```{r}
response <- "bottle_t_deg_c"
response
```
# Train Data
```{r}
sp <- group_initial_split(df_final, group = cast_sta_id, prop = 0.8)

sp
```

```{r}
train <- training(sp)

head(train)

```

```{r}
test  <- testing(sp)
head(test)
```
```{r}
# Keep Numeric Variables Only
train_num <- train %>%
  mutate(across(everything(),
                ~ suppressWarnings(as.numeric(.x))))
head(train_num)

```

```{r}
test_num <- test %>%
  mutate(across(everything(),
                ~ suppressWarnings(as.numeric(.x))))
head(test_num)
```

```{r}
train_num_d <- train_num %>% dplyr::select(where(is.numeric))
nrow(train_num)
nrow(train_num_d)
```
```{r}
test_num_d <- test_num %>% dplyr::select(where(is.numeric))
nrow(test_num)
nrow(test_num_d)
```


```{r}
dim(train_num_d)
summary(train_num_d$bottle_t_deg_c)
```

```{r}
sum(complete.cases(train_num_d))
```
Remove Very Sparse Columns

```{r}
#Calculate the missing data rate

missing_rate <- colMeans(is.na(train_num_d))

head(missing_rate)
```
```{r}
#Keep Columns With <80% Missing
good_cols <- names(missing_rate[missing_rate < 0.8])

train_filtered <- train_num_d[, good_cols]
test_filtered  <- test_num_d[, good_cols]

dim(train_filtered)
```
```{r}
# Remove Rows With Missing Values
train_clean <- train_filtered[complete.cases(train_filtered), ]
test_clean  <- test_filtered[complete.cases(test_filtered), ]

nrow(train_clean)
nrow(test_clean)
```
# Correlation Analysis
```{r}
numeric_vars <- train_clean %>%
  dplyr::select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")

#  Save Full Correlation Matrix
write.csv(
  cor_matrix,
  "calcofi_full_correlation_matrix.csv",
  row.names = TRUE
)

# Extract correlations with temperature
temp_cor <- cor_matrix["bottle_t_deg_c", ]

# Convert to tidy table
cor_table <- tibble(
  variable = names(temp_cor),
  correlation_with_temperature = round(as.numeric(temp_cor), 3)
) %>%
  arrange(desc(abs(correlation_with_temperature)))

# Save Temperature Correlation Table

write.csv(
  cor_table,
  "calcofi_temperature_correlations.csv",
  row.names = FALSE
)

cor_table
```
```{r}
getwd()
```

```{r}
#Select Only Strong Predictors

selected_vars <- c(
  "bottle_depthm",
  "bottle_salnty",
  "bottle_po4u_m",
  "bottle_no3u_m",
  "bottle_si_o3u_m",
  "bottle_o2sat"
)

train_selected <- train_clean[, c(response, selected_vars)]
test_selected  <- test_clean[, c(response, selected_vars)]
```

```{r}
#Remove Missing Rows

train_selected <- na.omit(train_selected)
test_selected  <- na.omit(test_selected)
```



```{r}
#Linear Regression
m_lm <- lm(bottle_t_deg_c ~ ., data = train_selected)

#pred_lm <- predict(m_lm, newdata = test_selected)

summary(m_lm)
```

```{r}
# Generate a correlation matrix for your predictors
cor_matrix <- cor(train_selected[, selected_vars])

# Save it as a CSV file
write.csv(cor_matrix, "correlation_matrix.csv")

round(cor_matrix, 2) # Rounding to 2 decimals 
```
```{r}
# Define your website folder path
website_folder <- "C:/Users/afsan/OneDrive/Desktop/Thesis/calcofi_dashboards/"

```


```{r}
# Create the heatmap
ggcorrplot(cor_matrix, 
           hc.order = TRUE,         # Groups similar variables together
           type = "lower",          # Shows only the bottom half
           lab = TRUE,              # Adds the actual numbers to the squares
           lab_size = 3,            # Size of the numbers
           colors = c("#E46726", "white", "#6D9EC1"), # Red for negative, Blue for positive
           title = "Correlation Heatmap of Ocean Predictors",
           ggtheme = theme_minimal())

ggsave(paste0(website_folder, "img-5.jpeg"), width = 8, height = 6, dpi = 300)
```


```{r}
#Ridge Regression
#  Prepare the data (X = predictors, y = target variable, temperature)
x <- as.matrix(train_selected[, selected_vars])
y <- train_selected$bottle_t_deg_c 

# 2. Find the best penalty value (lambda) using cross-validation
cv_model <- cv.glmnet(x, y, alpha = 0) # alpha=0 means Ridge
best_lambda <- cv_model$lambda.min

# 3. Fit the final Ridge model
ridge_model <- glmnet(x, y, alpha = 0, lambda = best_lambda)

# 4. See which ocean variables are most important after shrinkage
coef(ridge_model)


```
```{r}
# Linear Regression Prediction
# Using the test data model
pred_lm <- predict(m_lm, newdata = test_selected)

# Ridge Regression Prediction
# Ridge needs the data as a matrix
x_test <- as.matrix(test_selected[, selected_vars])
pred_ridge <- as.numeric(predict(ridge_model, newx = x_test, s = best_lambda))

# The Comparison 
results_comparison <- data.frame(
  Actual = test_selected$bottle_t_deg_c,
  Linear_Error = abs(test_selected$bottle_t_deg_c - pred_lm),
  Ridge_Error  = abs(test_selected$bottle_t_deg_c - pred_ridge)
)

# Calculate Average Error (MAE)
# If Ridge_MAE is lower, Ridge is officially better!
summary_metrics <- results_comparison %>%
  summarise(
    Linear_MAE = mean(Linear_Error),
    Ridge_MAE  = mean(Ridge_Error)
  )

print(summary_metrics)

```




```{r}
# 1. Get the model's predictions the data
predictions <- predict(ridge_model, s = best_lambda, newx = x)

# 2. Calculate the Sum of Squares (The math for R-squared)
actual <- y
rss <- sum((predictions - actual)^2)  # Residual Sum of Squares
tss <- sum((actual - mean(actual))^2) # Total Sum of Squares

# 3. The Final R-squared
r_squared <- 1 - (rss / tss)
print(paste("The R-squared value is:", round(r_squared, 4)))

```

```{r}
# Prepare the data
plot_lm_final <- data.frame(
  Actual = test_selected$bottle_t_deg_c,
  Predicted = as.numeric(pred_lm)
)

# Create the Plot
ggplot(plot_lm_final, aes(x = Actual, y = Predicted)) +
  # Points
  geom_point(alpha = 0.3, color = "#6D9EC1", size = 1.2) +
  
  # The 1:1 Identity Line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  
  # This allows the plot to stretch and fill the RStudio window
  scale_x_continuous(limits = c(5, 25)) +
  scale_y_continuous(limits = c(5, 25)) +
  
  # Better label placement (Top-Left)
  annotate("text", x = 5.5, y = 24, label = "Identity Line (1:1)", 
           color = "red", fontface = "italic", size = 4, hjust = 0) +
  
  #  Wrap title so it doesn't cut off
  labs(
    title = "Linear Regression Analysis:\nActual vs. Predicted Temperature",
    subtitle = "Model Accuracy (Adjusted R\u00B2): 0.8562",
    x = "Actual Temperature (\u00B0C)",
    y = "Predicted Temperature (\u00B0C)"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, lineheight = 1.2),
    plot.subtitle = element_text(size = 12, color = "darkgray"),
    axis.title = element_text(size = 11)
  )

ggsave(paste0(website_folder, "img-6.jpeg"), width = 8, height = 6, dpi = 300)

```


```{r}

# 1. Ensure we use the cleaned data (removing those 100°C errors)
ocean_data_filtered <- train %>%
  mutate(
    temp = as.numeric(bottle_t_deg_c),
    depth = as.numeric(bottle_depthm)
  ) %>%
  filter(temp < 35, !is.na(temp), !is.na(depth))

# 2. Plot Depth vs. Temperature with a Linear Regression line
ggplot(ocean_data_filtered, aes(x = depth, y = temp)) +
  # Add the raw data points (using alpha to handle the thousands of dots)
  geom_point(alpha = 0.2, color = "steelblue", size = 1) +
  # Add the Linear Regression Line in Red
  geom_smooth(method = "lm", color = "red", size = 1.2) + 
  labs(title = "Linear Regression: How Temperature Changes with Depth",
       subtitle = "The red line shows the predicted decrease in temperature",
       x = "Depth (m)", 
       y = "Temperature") +
  theme_minimal()

ggsave(paste0(website_folder, "img-7.jpeg"), width = 8, height = 6, dpi = 300)

```

```{r}
# 1. Prepare the data with a "Depth sizes"
ocean_data_faceted <- train %>%
  mutate(
    temp = as.numeric(bottle_t_deg_c),
    sal = as.numeric(bottle_salnty),
    depth = as.numeric(bottle_depthm),
    # Create a category to split the data at 200 meters
    depth_group = ifelse(depth <= 200, "Shallow Water (0-200m)", "Deep Water (>200m)")
  ) %>%
  # Filter out errors and missing values
  filter(temp < 35, !is.na(temp), !is.na(sal), !is.na(depth_group))

# 2. Create the Faceted Plot
ggplot(ocean_data_faceted, aes(x = temp, y = sal, color = depth)) +
  geom_point(alpha = 0.2, size = 0.8) +
  # Add a trend line to EACH panel
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  # This line creates the two side-by-side windows
  facet_wrap(~ depth_group, scales = "free_x") + 
  scale_color_viridis_c(direction = -1, name = "Depth (m)") +
  labs(title = "Temperature vs. Salinity by Depth Layer",
       subtitle = "Does the relationship flip in deep water?",
       x = "Temperature", 
       y = "Salinity (psu)") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 12)) # Make panel titles bold

ggsave(paste0(website_folder, "img-8.jpeg"), width = 10, height = 5, dpi = 300)

```

#Categorical Analysis


```{r}
# 1. Prepare the categorical data
depth_summary <- train %>%
  mutate(
    temp = as.numeric(bottle_t_deg_c),
    depth_group = ifelse(as.numeric(bottle_depthm) <= 200, "Shallow (0-200m)", "Deep (>200m)")
  ) %>%
  filter(!is.na(temp), !is.na(depth_group), temp < 40) %>%
  group_by(depth_group) %>%
  summarise(
    mean_temp = mean(temp),
    se = sd(temp) / sqrt(n()) # Standard Error for the bars
  )

# 2. Create the Bar Chart
ggplot(depth_summary, aes(x = depth_group, y = mean_temp, fill = depth_group)) +
  geom_bar(stat = "identity", color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_temp - se, ymax = mean_temp + se), width = 0.2) +
  scale_fill_manual(values = c("Deep (>200m)" = "#2c7fb8", "Shallow (0-200m)" = "#fec44f")) +
  labs(title = "Categorical Analysis: Average Temperature by Depth Layer",
       subtitle = "Comparing the sun-lit surface to the dark deep ocean",
       x = "Depth Category", 
       y = "Average Temperature (C)") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(paste0(website_folder, "img-9.jpeg"), width = 7, height = 5, dpi = 300)

```


Anova

```{r}
# 1. Prepare the data for the test
anova_data <- train %>%
  mutate(
    temp = as.numeric(bottle_t_deg_c),
    depth_group = ifelse(as.numeric(bottle_depthm) <= 200, "Shallow", "Deep")
  ) %>%
  filter(!is.na(temp), !is.na(depth_group))

# 2. Run the ANOVA (Comparing Temperature by Depth Group)
temp_anova <- aov(temp ~ depth_group, data = anova_data)

# 3. View the results
summary(temp_anova)

```






```{r}
# This lists only column names that contain "lat" or "lon"
colnames(train)[grep("lat|lon", colnames(train), ignore.case = TRUE)]

```




```{r}

library(leaflet)


# 1. Clean the data and prepare for the map
map_data <- train %>%
  mutate(
    lat = as.numeric(cast_lat_dec),
    lon = as.numeric(cast_lon_dec),
    temp = as.numeric(bottle_t_deg_c)
  ) %>%
  # Filter out missing coordinates and extreme temperature outliers
  filter(!is.na(lat), !is.na(lon), !is.na(temp), temp < 40) %>%
  group_by(cast_sta_id, lat, lon) %>%
  summarise(avg_temp = mean(temp), .groups = "drop")

# 2. Create a color palette (Blue for cold, Red for hot)
pal <- colorNumeric(palette = "YlOrRd", domain = map_data$avg_temp)

# 3. Build the Leaflet map
leaflet(map_data) %>%
  addTiles() %>%  # Adds the standard street map background
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius = 5,
    color = ~pal(avg_temp),
    stroke = FALSE, fillOpacity = 0.8,
    # This popup appears when you click a dot
    popup = ~paste0("<b>Station ID:</b> ", cast_sta_id, 
                   "<br><b>Avg Temp:</b> ", round(avg_temp, 2), "\u00B0C")
  ) %>%
  addLegend(pal = pal, values = ~avg_temp, title = "Temp (\u00B0C)")

```


```{r}
# Time Seasonal
# 1. Create the Season column and filter for depth layers
seasonal_data <- train %>%
  mutate(
    # Convert types for calculation
    temp = as.numeric(bottle_t_deg_c),
    sal = as.numeric(bottle_salnty),
    month = as.numeric(cast_month),
    depth = as.numeric(bottle_depthm),
    # Assign seasons based on month number
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ),
    depth_group = ifelse(depth <= 200, "Shallow (0-200m)", "Deep (>200m)")
  ) %>%
  filter(!is.na(season), !is.na(temp), !is.na(sal))

# 2. Create the Visual Comparison
ggplot(seasonal_data, aes(x = temp, y = sal, color = season)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) + 
  facet_wrap(~ depth_group, scales = "free") +
  labs(title = "Seasonal Impact on Temperature vs. Salinity",
       subtitle = "Does the time of year change the 'rules' of the ocean?",
       x = "Temperature (C)", y = "Salinity (psu)") +
  theme_minimal()

ggsave(paste0(website_folder, "img-11.jpeg"), width = 10, height = 5, dpi = 300)

```



```{r}
library(jsonlite)
write_json(map_data, "map_data.json", pretty = TRUE)


```
```{r}
head(map_data)
```
```{r}
library(jsonlite)
library(sf)
library(tidyverse)
library(janitor)

# 1. Load your JSON
# Using the path from your error message
path <- "C:/Users/afsan/OneDrive/Desktop/Thesis/map_data.json"
raw_data <- fromJSON(path) %>% as.data.frame()

# 2. Clean names to ensure consistency (janitor package)
clean_data <- raw_data %>% clean_names()

# 3. Identify coordinate columns automatically
# This looks for columns containing 'lat' or 'lon' / 'lng'
lon_col <- names(clean_data)[grepl("lon|lng", names(clean_data))][1]
lat_col <- names(clean_data)[grepl("lat", names(clean_data))][1]

# 4. Convert to spatial object using the auto-detected names
map_sf <- st_as_sf(clean_data, 
                   coords = c(lon_col, lat_col), 
                   crs = 4326)

# 5. Test the Map
library(leaflet)
leaflet(map_sf) %>%
  addTiles() %>%
  addCircleMarkers(radius = 3, color = "blue", weight = 1)
```




